### 2016-02-03 doc/thesis/intro.tex
Wrote about tree shapes

The aim of viral phylodynamics is to glean some kind of knowledge, about the
epidemic, the virus, or its hosts and their behaviour, by studying a phylogeny,
most often a transmission tree. Phylogenies are complex objects, and it is not
immediately obvious how to extract useful information from them with respect to
fitting a parameter. Standard statistical methods, built for numeric data
cannot be applied directly - for example, one cannot perform a regression of a
parameter of interest against a phylogeny. Therefore, before we discuss exactly
what phylodynamics can tell us about epidemics, and how it has been applied in
the past, we will review some of the methods for quantifying the shapes of
phylogenies and their similarity to each other.

Many tree summary statistics have been developed to assign numerical values to
phylogenies based on their properties. One of the most widely used is Sackin's
index~\cite{shao1990tree}, which measures the imbalance or asymmetry in the
tree.

An alternative to summary statistics are \defn{distance measures} on trees.
Rather than assigning a numerical value to each tree individually, a distance
measure assocates each pair of trees with a number, indicating how different
the trees are from each other. Distance measures allow us to identify groups of
related phylogenies, for example, local epidemics which are undergoing a
similar pattern of expansion.

@@

>>>>

The aim of viral phylodynamics is to glean some kind of knowledge, about the
epidemic, the virus, or its hosts and their behaviour, by studying a phylogeny,
most often a transmission tree~\autocite{pybus2009evolutionary, volz2013viral}.
Phylogenies are complex objects, and it is not immediately obvious how to
extract useful information from them with respect to fitting a parameter.
Standard statistical methods built for numeric data cannot be applied directly
- for example, one cannot perform a regression of a parameter of interest
against a phylogeny. Therefore, before we discuss exactly what phylodynamics
can tell us about epidemics, and how it has been applied in the past, we will
review some of the methods for quantifying the shapes of phylogenies and their
similarity to each other.

Many tree summary statistics have been developed to assign numerical values to
phylogenies based on their properties. One of the most widely used is Sackin's
index~\cite{shao1990tree}, which measures the imbalance or asymmetry in a
rooted tree. For the $i$th tip of the tree, we define $N_i$ to be the number of
branches between that tip and the root. The unnormalized Sackin's index is
defined as the sum of all $N_i$. It is called unnormalized because it does not
account for the number of tips in the tree. Among two trees having the same
number of tips, the least-balanced tree will have the highest Sackin's index.
However, among two equally balanced trees, the larger tree will have a higher
Sackin's index. This makes it challenging to compare balances among trees of
different sizes. To correct this, \textcite{kirkpatrick1993searching} derive
the expected value of Sackin' index under the Yule
model~\autocite{yule1925mathematical}. Dividing by this expected value
normalizes Sackin's index, so that it can be used to compare trees of different
sizes.

An alternative to summary statistics are \defn{distance measures} on trees.
Rather than assigning a numerical value to each tree individually, a distance
measure assocates each pair of trees with a number, indicating how different
the trees are from each other. Distance measures allow us to identify groups of
related phylogenies, for example, local epidemics which are undergoing a
similar pattern of expansion. One such distance measure is the
\gls{nltt}~\autocite{janzen2015approximate}, which compares the
\gls{ltt}~\autocite{nee1992tempo} plots of two trees. Specifically, the two
\gls{ltt} plots are normalized so that they begin at $(0, 0)$ and end at $(1,
1)$, and the difference between the two plots is integrated between 0 and 1.
In the context of infectious diseases, the \gls{nltt} is related to the
prevalence~\autocite{holmes1995revealing}, so large values may indicate that
the trees being compared are the products of different epidemic
trajectories~\autocite{janzen2015approximate}.

Another tree distance measure is the \defn{phylogenetic kernel}, or ``tree
kernel'' developed by \textcite{poon2013mapping}. Strictly speaking, the tree
kernel is a \defn{similarity measure}, rather than a distance measure, as it is
maximized when the two trees being compared are the same. The basis of the tree
kernel is the kernel trick originally developed for
\glspl{SVM}~\autocite{burges1998tutorial}. The idea of the kernel trick is to
compare objects by mapping them into a feature space of very high, possibly
even infinite, dimension. The similarity between objects is taken to be their
dot product in the feature space. It is called a ``trick'' because this dot
product is computed using a \defn{kernel function} without explicitly mapping
the objects to the feature space, which would be computationally prohibtive. In
the case of the tree kernel, the feature space is the space of all possible
\defn{subset trees}, which are subtrees that do not necessarily extend all the
way to the tips. The subset-tree kernel was originally developed for comparing
parse trees in natural language processing~\autocite{collins2002new} and did
not incorporate branch length information. The version developed by
\textcite{poon2013mapping} includes a radial basis function to compare the
differences in branch lengths, thus incorporating both the trees' topologies
and their branch lengths in a single similarity score. The tree kernel was
later shown to be highly effective in differentiating trees simulated under a
compartmental model with two risk groups of varying contact
rates~\autocite{poon2015phylodynamic}. In that paper,
\citeauthor{poon2015phylodynamic} used the tree kernel as the distance function
in \gls{ABC} (see Section~\ref{sec:abc}), an approach dubbed kernel-\gls{ABC},
to fit epidemiological models to observed trees.

@@

\label{sec:abc}
